### 2013-06-25 Tex/chapters/introduction.tex
Ausblick: Code bei Github
Anmerkung zu k_B = 1 nach vorner verschoben
Satyumstellungen und Korrekturen

The computing capabilities of modern computers enable researchers to
collect and analyse vast amounts of data.
Further statistical systems, for which exist no analytical soultions
or only for very simplified or special cases, can be simulated.
%~ Simulations also have the advantage to be able to measure some quantities
%~ which are not obtainable by experiment.
For example to examine phasetransitions, which are defined by the abrupt
change of an observable i.e. the change of the magnetisation of a
ferromagnet near the Curie temperature or the change of the density of
water near boiling. Unfortunatly the memory of any computer is small in
comparison with the thermodynamic limit.
%~ Of course it is only possible to simulate systems of finite size because
%~ of the finite memory of computers so the thermodynamic limit can not be
%~ simulated directly.
This leads to \emph{finite size effects}. In section \ref{ssec:finitesize}
will be discussed how to manage them.\\

One of the simplest models with a second order phasetransition is the
Ising model \cite{Ising1925} which is a simple model of a ferromagnet
and will be explained in more detail in section \ref{ssec:isingmodel}.
It is analyticaly solved for two dimensions on some regular lattices
\cite{Onsager1944} \cite{Wannier1945}.
In this thesis the critical behavior on some irregular lattices
corresponding to proximity graphs (see section \ref{ssec:graphtypes})
will be examined using the Monte Carlo simulations described in section
\ref{sec:montecarlo}.

>>>>

The computing capabilities of modern computers enable researchers to
collect and analyse vast amounts of data.
Further statistical systems, for which exist no analytical soultions
or only for very simplified or special cases, can be simulated.
%~ Simulations also have the advantage to be able to measure some quantities
%~ which are not obtainable by experiment.
For example to examine phasetransitions, which are defined by the abrupt
change of an observable i.e. the change of the density of water near
boiling or the change of the magnetisation of a ferromagnet near the
Curie temperature. Unfortunatly the memory of any computer is small in
comparison with the thermodynamic limit.
%~ Of course it is only possible to simulate systems of finite size because
%~ of the finite memory of computers so the thermodynamic limit can not be
%~ simulated directly.
This leads to \emph{finite size effects}. In section \ref{ssec:finitesize}
will be discussed how to manage them.\\
One of the simplest models with a second order phasetransition is the
Ising model \cite{Ising1925} which is a simple model of a ferromagnet
and will be explained in more detail in section \ref{ssec:isingmodel}.
It is analytically solved for two dimensions on some regular lattices
\cite{Onsager1944} \cite{Wannier1945}.
In this thesis the critical behavior on some irregular lattices
corresponding to proximity graphs (see section \ref{ssec:graphtypes})
will be examined using the Monte Carlo simulations described in section
\ref{sec:montecarlo}. Note that in the scope of this thesis
the Boltzmann constant \(k_{B}=1\) for the sake of simplicity.
All source code is available at \url{github [noch hochzuladen]}.

### 2013-06-25 Tex/chapters/model.tex
Satyumstellungen und Korrekturen

%~ \subsubsection{Graphs}
    %~ A graph is a set of nodes \(V\) and edges \(E\).
\subsection{Gabriel- and Relative Neighborhood Graph}
\label{ssec:graphtypes}
    A graph \(G(V,E)\) is a set of nodes \(V\) and edges \(E\).\\
    All here mentioned graph types are \emph{proximity graphs}. They are
    connecting nodes which are by some metric near to each other.
    Hence they are suited to generalize problems defined on regular
    lattices with nearest neighbor relationships, like the Ising model
    which will be described in more detail in section
    \ref{ssec:isingmodel}.
    In this thesis the distance is always determined by the Euclidian
    metric, though in principle every metric can be used.\\

    The Gabriel graph \cite{Gabriel1969} is a subgraph of the
    Delaunay triangulation. Two nodes \(i\) and \(j\) with distance
    \(d_{ij}\) are connected with an edge if a circle with it's
    center on the middlepoint between \(i\) and \(j\) and radius
    \(r = \frac d 2\) contains no other nodes. This area will be
    called \emph{lune} in the following. See also Figure
    \ref{fig:lunes}\subref{sfig:lunes:def}.\\
    The Relative Neighborhood graph \cite{Toussaint1980} is a
    subgraph of the Gabriel graph. Two nodes \(i\) and \(j\) with
    distance \(d_{ij}\) are connected if no other node is in the
    \emph{lune}. The lune is defined as the intersection of two
    circles with radius \(r = d\) and centers on \(i\) and \(j\).
    See also Figure \ref{fig:lunes}\subref{sfig:lunes:def}.
    \begin{figure}[htbp]
        \centering
        \subfigure[Definition of the lunes][]{
            \label{sfig:lunes:def}
            \input{images/GG_RNG_def}
        }
        \subfigure[Relative Neighborhood graph example][]{
            \label{sfig:lunes:rng}
            \includegraphics[width=0.3\textwidth]{images/RNG/L12S03.pdf}
        }
        \subfigure[Gabriel graph example][]{
            \label{sfig:lunes:gg}
            \includegraphics[width=0.3\textwidth]{images/GG/L12S03.pdf}
        }
        \caption[Gabriel - and Relative Neighborhood Graph]
        {
            \subref{sfig:lunes:def} Lunes of Relative Neighborhood
                Graph (hatched region) and
                Gabriel Graph (cross hatched region)
            \subref{sfig:lunes:rng} Example of a Relative
                Neighborhood Graph on periodic boundary conditions.
                Periodic nodes are dashed.
            \subref{sfig:lunes:gg} Example of a Gabriel Graph on
                periodic boundary conditions. Periodic nodes are
                dashed.
        }
        \label{fig:lunes}
    \end{figure}\\
    To construct these graphs the simple way is to test for each
    pair of nodes one has to check for every other node if it lies in
    the lune of the pair. That is of complexity \(O (N^3)\), because
    there are \(N(N-1)\) pairs and for each \(N-2\) nodes to test. So
    the product is of order \(O(N^3)\)\\
    To reduce the complexity one can first create a Delaunay
    Triangulation in complexity \(O (N \log N)\)
    \cite{Leach1992} and test the criterion for each edge, because
    the Delaunay triangulation is a supergraph of both. But the
    implementation of a Delaunay triangulation algorithm is not trivial
    and the generation of the graphs is not time critical in the scope
    of this bachelor thesis.\\
    So a tradeoff is to use basicly the simple method but only test
    the criterion for nodes which are near to the lune and abort if
    one node inside the lune is found. To determine which nodes are
    near the lune one can subdivide the area in \emph{cells} and save
    for each cell a list with nodes lying inside it like presented in
    \cite{RNGCell}.
    Now it is just neccessary to test the nodes in the cells which
    resemble a rectangular bounding box of the lune. Most pairs will be
    far away from each other and the cells in the middle of the bounding
    box are completely inside the lune so that only one node has to be
    tested to discard an edge between them. Connected nodes are near to
    each other so that only very few cells have to be tested.\\
    Indeed this method reduced the time needed to construct a Relative
    Neighborhood graph with \(N=32^2\) and \(N=64^2\) by a factor of
    over \(15\) respectivly \(40\). Though the complexity is still of
    order \(O(N^2)\) in the best case, because for every pair at least
    one check has to be performed.

\subsection{The Disordered Ising Model}
\label{ssec:isingmodel}
    The examined model is a modified 2D Ising model where the \(N=L^2\)
    sites are displaced nodes of a square lattice of edge length \(L\)
    with periodic boundary conditions. The displacement is randomly gauss
    distributed with the standard deviation \(\sigma\). This \(\sigma\)
    is also called \emph{disorder paramter} in the following.
    \begin{equation}
        \hat{H} = \sum_{\avg{i,j}}J_{ij}s_{i}s_{j} - H \sum_i s_i
        \label{eq:hamiltonian}
    \end{equation}
    Equation \eqref{eq:hamiltonian} is the hamiltonian of the Ising model.
    The magnetic field \(H=0\) in the scope of this thesis.
    \(\avg{i,j}\) refers to nodes \(i\) and \(j\) connected by an edge \(E_{ij}\).
    The edges are constructed according to one of the two the in section
    \ref{ssec:graphtypes} defined rules. So that the lattice represents
    a proximity graph. Each node \(i\) has a spin \(s_i = \pm 1\). Each
    edge \(E_{ij}\) has a weight \(J_{ij} = \exp (\alpha (1-d_{ij}))\).
    \(d_{ij}\) is the Euclidian distance between the nodes \(i\) and
    \(j\) and \(\alpha = 0.5\). \(J\) is called \emph{coupling constant}.\\
    For \(\sigma = 0\) this is the standard Ising model with \(J = 1\),
    for which exists an analytical solution \cite{Onsager1944}.\\

>>>>

%~ \subsubsection{Graphs}
    %~ A graph is a set of nodes \(V\) and edges \(E\).
\subsection{Gabriel- and Relative Neighborhood Graph}
\label{ssec:graphtypes}
    A graph \(G(V,E)\) is a set of nodes \(V\) and edges \(E\).\\
    All here mentioned graph types are \emph{proximity graphs}. They are
    connecting nodes which are by some metric near to each other.
    Hence they are suited to generalize problems defined on regular
    lattices with nearest neighbor relationships, like the Ising model
    which will be described in more detail in section
    \ref{ssec:isingmodel}.
    In this thesis the distance is always determined by the Euclidian
    metric in two dimensions, though in principle every metric in any
    dimension can be used.\\

    The Gabriel graph \cite{Gabriel1969} is a subgraph of the
    Delaunay triangulation. Two nodes \(i\) and \(j\) with distance
    \(d_{ij}\) are connected with an edge, if a circle with it's
    center on the middlepoint between \(i\) and \(j\) and radius
    \(r = \frac d 2\) contains no other nodes. This area will be
    called \emph{lune} in the following. See also Figure
    \ref{fig:lunes}\subref{sfig:lunes:def}.\\
    The Relative Neighborhood graph \cite{Toussaint1980} is a
    subgraph of the Gabriel graph. Two nodes \(i\) and \(j\) with
    distance \(d_{ij}\) are connected, if no other node is in the
    \emph{lune}. The lune is defined as the intersection of two
    circles with radius \(r = d\) and centers on \(i\) and \(j\).
    See also Figure \ref{fig:lunes}\subref{sfig:lunes:def}.
    \begin{figure}[htbp]
        \centering
        \subfigure[Definition of the lunes][]{
            \label{sfig:lunes:def}
            \input{images/GG_RNG_def}
        }
        \subfigure[Relative Neighborhood graph example][]{
            \label{sfig:lunes:rng}
            \includegraphics[width=0.3\textwidth]{images/RNG/L12S03.pdf}
        }
        \subfigure[Gabriel graph example][]{
            \label{sfig:lunes:gg}
            \includegraphics[width=0.3\textwidth]{images/GG/L12S03.pdf}
        }
        \caption[Gabriel - and Relative Neighborhood Graph]
        {
            \subref{sfig:lunes:def} Lunes of Relative Neighborhood
                Graph (hatched region) and
                Gabriel Graph (cross hatched region)
            \subref{sfig:lunes:rng} Example of a Relative
                Neighborhood Graph on periodic boundary conditions.
                Periodic nodes are dashed.
            \subref{sfig:lunes:gg} Example of a Gabriel Graph on
                periodic boundary conditions. Periodic nodes are
                dashed.
        }
        \label{fig:lunes}
    \end{figure}\\
    To construct these graphs the simple way is to test for each
    pair of nodes one has to check for every other node if it lies in
    the lune of the pair. That is of complexity \(O (N^3)\), because
    there are \(N(N-1)\) pairs and for each \(N-2\) nodes to test. So
    the product is of order \(O(N^3)\)\\
    To reduce the complexity one can first create a Delaunay
    Triangulation in complexity \(O (N \log N)\)
    \cite{Leach1992} and test the criterion for each edge, because
    the Delaunay triangulation is a supergraph of both. But the
    implementation of a Delaunay triangulation algorithm is not trivial
    and the generation of the graphs is not time critical in the scope
    of this bachelor thesis.\\
    So a tradeoff is to use basicly the simple method but only test
    the criterion for nodes which are near to the lune and abort if
    one node inside the lune is found. To determine which nodes are
    near the lune one can subdivide the area in \emph{cells} and save
    for each cell a list with nodes lying inside it like presented in
    \cite{RNGCell}.
    Now it is just neccessary to test the nodes in the cells which
    resemble a rectangular bounding box of the lune. Most pairs will be
    far away from each other and the cells in the middle of the bounding
    box are completely inside the lune so that only one node has to be
    tested to discard an edge between them. Connected nodes are near to
    each other so that only very few cells have to be tested.\\
    Indeed this method reduced the time needed to construct a Relative
    Neighborhood graph with \(N=32^2\) and \(N=64^2\) by a factor of
    over \(15\) respectivly \(40\). Though the complexity is still of
    order \(O(N^2)\) in the best case, because for every pair at least
    one check has to be performed.

\subsection{The Disordered Ising Model}
\label{ssec:isingmodel}
    The examined model is a modified 2D Ising model where the \(N=L^2\)
    sites are displaced nodes of a square lattice of edge length \(L\)
    with periodic boundary conditions. The displacement is randomly gauss
    distributed with the standard deviation \(\sigma\). This \(\sigma\)
    is also called \emph{disorder paramter} in the following.
    \begin{equation}
        \hat{H} = \sum_{\avg{i,j}}J_{ij}s_{i}s_{j} - H \sum_i s_i
        \label{eq:hamiltonian}
    \end{equation}
    Equation \eqref{eq:hamiltonian} is the hamiltonian of the Ising model.
    The magnetic field \(H=0\) in the scope of this thesis.
    \(\avg{i,j}\) refers to nodes \(i\) and \(j\) connected by an edge \(E_{ij}\).
    The edges are constructed according to one of the two the in section
    \ref{ssec:graphtypes} defined rules. So that the lattice represents
    a proximity graph. Each node \(i\) has a spin \(s_i = \pm 1\). Each
    edge \(E_{ij}\) has a weight \(J_{ij} = \exp (\alpha (1-d_{ij}))\).
    \(d_{ij}\) is the Euclidian distance between the nodes \(i\) and
    \(j\) and \(\alpha = 0.5\). \(J\) is called \emph{coupling constant}.
    This is definition is inspired by \cite{Lima2000}.\\
    For \(\sigma = 0\) this is the standard Ising model with \(J = 1\),
    for which exists an analytical solution \cite{Onsager1944}.\\

### 2013-06-25 Tex/chapters/monte_carlo.tex
Fehlerkorrekturen und Formulierungsverbesserungen
Satyumstellungen und Korrekturen

    \paragraph{Wolff}
        Close to the critical temperature \(T_c\) the Metropolis
        gets slower. This is called \emph{critical slowing} down and the
        cause is beyond the scope of this thesis.\\
        Using a cluster algorithm like the Wolff
        algorithm \cite{Wolff1989} speeds things up.
        For an Ising model the Wolff algorithm builds a cluster of sites
        starting with a random site and adding neighboring sites of the
        same spin with probability
        \(P_{\mathrm{add}} = 1-\exp\brac{-2\beta J}\)
        where \(J\) is the coupling constant (explained in section
        \ref{ssec:isingmodel}). The neighboring sites of the added sites
        are also considered and so forth. When there are no more sites
        to add, the spin of every site in the cluster is flipped
        \cite[S. ??]{NewmanBarkema1999} \cite[S. 151f]{Katzgraber2011}.
        This leads fast to new uncorrelated states at the critical
        temperature because big clusters are flipped. But there are not
        much advantages at high or low temperatures. At low temperatures
        the cluster will consist of almost all sites such that all but
        very few spins will be flipped. At high temperatures the cluster
        will only contain very few sites.
        Both situaitions have no advantage against the Metropolis algorithm.

@@

    \paragraph{Implementation Details}
        Here a mixture of the above three algorithms is used.
        Each sweep \(N\) Metropolis spin flips, one Wolff cluster flip
        and one Parallel Tempering swap are performed. Where \(N\) is the
        count of sites.
%~ Monte Carlo Simulationen:\\
    %~ Wolff-Cluster Algorithmus \cite{Wolff1989} (siehe auch \cite[S. xx]{NewmanBarkema1999}),
    %~ Metropolis Sweep \cite{Metropolis1953} (siehe auch \cite[S. xx]{NewmanBarkema1999}),
    %~ Parallel Tempering \cite{ParallelTempering1986} (siehe auch \cite[S. xx]{NewmanBarkema1999} \cite[S. xx]{Katzgraber2011})\\

>>>>

    \paragraph{Wolff}
        Close to the critical temperature \(T_c\) the Metropolis
        gets slower. This is called \emph{critical slowing} down and the
        cause is beyond the scope of this thesis.\\
        Using a cluster algorithm like the Wolff
        algorithm \cite{Wolff1989} speeds things up.
        For an Ising model the Wolff algorithm builds a cluster of sites
        starting with a random site and adding neighboring sites of the
        same spin with probability \(P_{\mathrm{add}}\) from eq. \eqref{eq:wolffAdd}
        \begin{equation}
            P_{\mathrm{add}} = 1-\exp\brac{-2\beta J}
            \label{eq:wolffAdd}
        \end{equation}
        Where \(J\) is the coupling constant (explained in section
        \ref{ssec:isingmodel}). The neighboring sites of the added sites
        are also considered and so forth. When there are no more sites
        to add, the spin of every site in the cluster is flipped
        \cite[S. ??]{NewmanBarkema1999} \cite[S. 151f]{Katzgraber2011}.
        This leads fast to new uncorrelated states at the critical
        temperature because big clusters are flipped. But there are not
        much advantages at high or low temperatures. At low temperatures
        the cluster will consist of almost all sites such that all but
        very few spins will be flipped. At high temperatures the cluster
        will only contain very few sites.
        Both situaitions have no advantage against the Metropolis algorithm.

@@

    \paragraph{Implementation Details}
        Here a mixture of the above three algorithms is used.
        Each sweep \(N\) Metropolis spin flips, one Wolff cluster flip
        and one Parallel Tempering swap are performed. Where \(N\) is the
        count of sites. All needed random numbers are generated by the
        GSL \cite{GSL} implementation of the \emph{Mersenne Twister} \cite{Matsumoto1998}.
%~ Monte Carlo Simulationen:\\
    %~ Wolff-Cluster Algorithmus \cite{Wolff1989} (siehe auch \cite[S. xx]{NewmanBarkema1999}),
    %~ Metropolis Sweep \cite{Metropolis1953} (siehe auch \cite[S. xx]{NewmanBarkema1999}),
    %~ Parallel Tempering \cite{ParallelTempering1986} (siehe auch \cite[S. xx]{NewmanBarkema1999} \cite[S. xx]{Katzgraber2011})\\
